\pkg{evtree}: Evolutionary Learning of Globally Optimal Classification and Regression Trees in \proglang{R}

### 2011-09-28 pkg/evtree/inst/doc/evtree.Rnw
fixed FIXMEs
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@250 255f4249-2263-4417-b1ea-8498779e27a7

Instead of searching all combinations in $\Theta$ simultaneously, recursive
partitioning algorithms only consider one split at a time. At each internal node
$r \in \{n_1,..., n_{M-1}\}$, the split variable $v_r$ and the corresponding
split point $s_r$ are selected to locally minimize the loss function.
Starting with an empty tree $\theta_0 = (\emptyset)$, the tree is first grown 
recursively for and subsequently \emph{pruned} to satisfy the complexity tradeoff:
%
\begin{eqnarray} \label{eq:locally}
  \bar \theta_r
    & = & \argmin_{ \theta = \theta_{r-1} \cup (v_r, s_r)} \; \mathrm{loss}\{Y, f(X, \theta)\} 
    \qquad (r = 1, \dots, M_{\max} -1), \\
  \bar \theta
    & = & \argmin_{\bar \theta_r} \; \mathrm{loss}\{Y, f(X, \bar \theta_r)\} ~+~ \mathrm{comp}(\bar \theta_r).
  \label{eq:pruning}
\end{eqnarray}
%
For nontrivial problems, forward-search recursive partitioning methods only search a small
fraction of the global search space $(v_1, s_1, \dots, v_{M_{\max}-1},
s_{M_{\max}-1} )$. They only search each $(v_r, s_r)$ once, and independently of
the subsequent split rules, hence typically leading to a globally suboptimal solution
$\bar \theta$.

@@

When compared with the described forward stepwise search, a less greedy approach
is to calculate the effects of the split rules deeper down in the
tree. In this way optimal trees can be found for simple problems. However,
split selection at a given node in Equation~\ref{eq:locally} has complexity  $O(PN)$
(if all $P$ variables are numeric/ordered with $N$ distinct values).
Through a global search up to $M-1$ levels (\ref{eq:globally}) the complexity
increases to $O(P^M N^M)$ \citep{papagelis2001bdt}. One conceivable compromise
between these two extremes is to look ahead $d$~steps with $1 < d < M$
\citep[see e.g.,][]{esmeir2007ald}, also yielding a locally optimal tree but
less constrained than that from a 1-step-ahead search.
Another class of algorithms is given by stochastic optimization methods that,
given an initial tree, seek improved solutions through stochastic changes to
the tree structure. Thus, these algorithms try to explore the full parameter
space $\Theta$ but cannot be guaranteed to find the globally optimal solution
but only an approximation thereof. Besides evolutionary algorithms
\citep{koza1991concept}, \textit{Bayesian CART} \citep{denison1998bca} and
\textit{simulated annealing} \citep{sutton1992ict} were used successfully to
solve difficult classification and regression tree problems.
\cite{koza1991concept} first formulated the concept of using evolutionary
algorithms as a stochastic optimization method to build classification and
regression trees. \cite{papagelis2001bdt} presented a classification tree
algorithm and provided results on several datasets from the UCI machine learning
repository \citep{Frank+Asuncion:2010}. Another method for the construction of
classification and regression trees via evolutionary algorithms was introduced by
\cite{gray2008cta} and \cite{fan2005rta}, respectively.
\cite{cantu2003inducing} used an evolutionary algorithm to induce so-called oblique
classification trees.

@@

\fixme{I really had problems following this section. The sub-operator is useful
but is it possible to introduce this with fewer sub- and super-scripts? It makes
it very challenging to read.}

\textit{Crossover} exchanges, randomly selected, subtrees between two trees. Let
$\tilde{\theta}_{i}^1$ and $\tilde{\theta}_{i}^2$ be the parameters of the trees
that are chosen for crossover. First, an internal node is selected randomly from
each of $\tilde{\theta}_{i}^1$ and $\tilde{\theta}_{i}^2$, which we denote as
$n_k^1$ and $n_k^2$. Let $\mathrm{sub}1(\tilde{\theta}_i^j, n^j_k)$ denote the subtrees
rooted by $n_k^j$ ($j=1,2$). Then, $\mathrm{sub}1(\tilde{\theta}_i^j) =
(v_{n^j_k}, s_{n^j_k},  v_{n^j_{k1}}, s_{n^j_{k1}}, \dots ,v_{n^j_{km}}, s_{n^j_{km}})$, where $(n^j_{k1}, ..., n^j_{km})$ are the set
of descendent node positions of $n^j_k$. Now, the two selected subtrees can be
written as $\mathrm{sub}1(\tilde{\theta}_{i}^1, n^1_k) $ and $\mathrm{sub}1(\tilde{\theta}_{i}^2,
n^2_k)$. Furthermore, $\mathrm{sub}2(\tilde{\theta}_{i}^1, n^1_k) = \tilde{\theta}_{i}^1
\setminus \mathrm{sub}1(\tilde{\theta}_{i}^1, n^1_k)$ and $\mathrm{sub}2(\tilde{\theta}_{i}^2,
n^2_k) = \tilde{\theta}_{i}^2 \setminus \mathrm{sub}1(\tilde{\theta}_{i}^2, n^2_k)$.

After the application of the \textit{crossover} operator, the trees' parameter
spaces are given by $\tilde{\theta}_{i+1}^1= \mathrm{sub}2(\tilde{\theta}_{i}^1, n^1_k)
\cup \mathrm{sub}1(\tilde{\theta}_{i}^2, n^2_k) \setminus (v_{m^2_{1}}, s_{m^2_{1}}, ...,
v_{m^2_{f}}, s_{m^2_{f}})$ and $\tilde{\theta}_{i+1}^2=
\mathrm{sub}2(\tilde{\theta}_{i}^2, n^2_k) \cup \mathrm{sub}1(\tilde{\theta}_{i}^1, n^1_k)
\setminus (v_{m^1_{1}}, s_{m^1_{1}}, ..., v_{m^1_{f}}, s_{m^1_{f}})$. Here,
$(v_{m^1_{1}}, s_{m^1_{1}}, ..., v_{m^1_{f}}, s_{m^1_{f}}) \subset
\mathrm{sub}1(\tilde{\theta}_i^1, n^1_k)$ and $(v_{m^2_{1}}, s_{m^2_{1}}, ...,
v_{m^2_{f}}, s_{m^2_{f}}) \subset \mathrm{sub}1(\tilde{\theta}_i^2, n^2_k)$ are the
parameters of the invalid nodes.

@@

First, the data is loaded and the forward-search trees are grown
with and without depth restriction, visualizing the unrestricted trees
in Figure~\ref{fig:example_rpart_ctree}.
%
<<BBBClub-rpart-ctree, echo=TRUE, eval=FALSE>>=
data("BBBClub", package = "evtree")
library("rpart")
rp  <- as.party(rpart(choice ~ ., data = BBBClub, minbucket = 10))
rp2 <- as.party(rpart(choice ~ ., data = BBBClub, minbucket = 10,
  maxdepth = 2))
ct  <- ctree(choice ~ ., data = BBBClub, minbucket = 10, mincrit = 0.99)
ct2 <- ctree(choice ~ ., data = BBBClub, minbucket = 10, mincrit = 0.99,
  maxdepth = 2)
plot(rp)
plot(ct)
@
%
With the objective of building a smaller, but at still accurate tree,
\code{evtree} is constrained to a maximum tree depth of $2$,
see Figure~\ref{fig:example_evtree}.
%
<<BBBClub-evtree, echo=TRUE, eval=FALSE>>=
set.seed(1090)
ev <- evtree(choice ~ ., data = BBBClub, minbucket = 10, maxdepth = 2)
@
%
<<BBBClub-cache, echo=FALSE, eval=TRUE>>=
if(cache & file.exists("BBBClub-trees.rda")) {
load("BBBClub-trees.rda")
} else {
<<BBBClub-rpart-ctree>>
<<BBBClub-evtree>>
if(cache) {
  save(rp, rp2, ct, ct2, ev, file = "BBBClub-trees.rda")
} else {
  if(file.exists("BBBClub-trees.rda")) file.remove("BBBClub-trees.rda")
}
}
@
%
\setkeys{Gin}{width=1\textwidth}
\begin{figure}[p!]
\centering
\begin{minipage}{1\textwidth}
\begin{flushleft} \code{rpart} \end{flushleft}
\centering
\setkeys{Gin}{width=0.75\textwidth}
<<BBBClub-rpart-plot, fig=TRUE, echo=FALSE, width=8.1, height=7>>=
plot(rp)
@
\end{minipage}
\begin{minipage}{1\textwidth}
\begin{flushleft} \code{ctree} \end{flushleft}
\centering
\setkeys{Gin}{width=1\textwidth}
<<BBBClub-ctree-plot, fig=TRUE, echo=FALSE,  width=11.4, height=5.9>>=
plot(ct)
@
\end{minipage}
\caption{\label{fig:example_rpart_ctree} Trees for customer targeting constructed by \code{rpart}
  (upper panel) and \code{ctree} (lower panel). The target variable is the customer's choice of buying the book.
  The variables used for splitting are the number of art books purchased previously (\code{art}), the number
  of months since the first purchase (\code{first}), the frequency of previous purchases at the Bookbinder's Book Club
  (\code{freq}) and the customer's \code{gender}.}
\end{figure}
%
The resulting evolutionary tree is printed output below and visualized in
Figure~\ref{fig:example_evtree}.
%
<<BBBClub-evtree-display>>=
plot(ev)
ev
@
%
\setkeys{Gin}{width=1\textwidth}
\begin{figure}[t!]
\begin{flushleft} \code{evtree} \end{flushleft}
\centering
\setkeys{Gin}{width=0.7\textwidth}
<<BBBClub-evtree-plot, fig=TRUE, echo=FALSE, width=7.5, height=4.9>>=
plot(ev)
@
\caption{\label{fig:example_evtree} Tree for customer targeting constructed by \code{evtree}
   The target variable is the customer's choice of buying the book. The variables used for splitting are
   the number of art books purchased previously (\code{art}) and the number of months since the first purchase
    \code{first}.}
\end{figure}

@@

In the first part of the analysis (Section~\ref{comp1}) the algorithms are
compared on 14~benchmark datasets that are publicly available and $3$ real-world
datasets from the Austrian \textit{Diagnosis Related Group (DRG)} system
\citep{LKF2009}. The analysis is based on the evaluation of 250~bootstrap
samples for each of the 20~datasets. The misclassification rate on the
\textit{out-of-bag} \citep{hothorn2005design} samples is used as a measure of
predictive accuracy. Furthermore, the complexity is estimated by the number of
terminal nodes.

@@

In Table~\ref{tab:dataDescription} the benchmark and real-world datasets from
the Austrian DRG-system are described. 
\fixme{The DRG description is far too long. All readers are waiting for the results
of the benchmark now. One explanatory sentence -- or maybe two at most -- should be
used to state the general problem and point to the manual pages.}
In the Austrian DRG-system, resources are
allocated to hospitals by simple rules regarding diagnosis, selected individual
services, and patient characteristics. Within this system regression tree
analysis is performed to classify hospital patients into clinically meaningful
and comprehensible groups that consume similar hospital resources. The datasets
MEL0101 (intervention on the skull), HDG0202 (malignant neoplasms), and HDG0502
(acute affections of the respiratory tract and middle atelectasis) contain
patient groups with the same main diagnosis or procedure. The task of the
regression tree analysis is to further divide these groups with the intention of
finding groups with similar resource consumption. Possible split-variables are
the patients' diagnoses, procedures, age, and gender. The dependent variable is
the patients' length of stay, which has a good relationship with the total costs
and is available \citep{theurl2007ihf}. Permission to use the data was granted
by the \citet{BMG2009}. A more detailed description of the application can be
found in \citet{grubinger2010regression}.

@@

The disadvantages of the \code{evtree} algorithm are computation time and memory requirements. While the smallest of the analyzed datasets, \textit{Glass identification},
only needed approximately 4--6~seconds to fit, larger datasets demanded several minutes. The fit of a model from the largest dataset, \textit{MAGIC gamma telescope}, required approximately 40--50 minutes and a main memory of 400 Mbit. The required resources were measured on an Intel Core 2 Duo with 2.2 GHz and 2 GB RAM using the 64-bit version of Ubuntu 10.10.
\fixme{also briefly discuss the issue of instability of the evolutionary trees even for large populations and number of runs}

@@

In this paper, we presented the \pkg{evtree} package, which implements
classification and regression trees that are grown by through an evolutionary
algorithm. The package uses standard \code{print()}, \code{summary()}, and
\code{plot()} functions to display trees and a \code{predict()} function to
predict the class labels of new data from the \pkg{partykit} package. As evolutionary
learning of trees is computationally demanding, most calculations
are conducted in \proglang{C++}. At the moment our algorithm does not support
parallelism. However, we intend to extend \pkg{evtree} to parallel computing.

>>>>

Instead of searching all combinations in $\Theta$ simultaneously, recursive
partitioning algorithms only consider one split at a time. At each internal node
$r \in \{n_1,..., n_{M-1}\}$, the split variable $v_r$ and the corresponding
split point $s_r$ are selected to locally minimize the loss function.
Starting with an empty tree $\theta_0 = (\emptyset)$, the tree is first grown 
recursively and subsequently \emph{pruned} to satisfy the complexity tradeoff:
%
\begin{eqnarray} \label{eq:locally}
  \bar \theta_r
    & = & \argmin_{ \theta = \theta_{r-1} \cup (v_r, s_r)} \; \mathrm{loss}\{Y, f(X, \theta)\} 
    \qquad (r = 1, \dots, M_{\max} -1), \\
  \bar \theta
    & = & \argmin_{\bar \theta_r} \; \mathrm{loss}\{Y, f(X, \bar \theta_r)\} ~+~ \mathrm{comp}(\bar \theta_r).
  \label{eq:pruning}
\end{eqnarray}
%
For nontrivial problems, forward-search recursive partitioning methods only search a small
fraction of the global search space $(v_1, s_1, \dots, v_{M_{\max}-1},
s_{M_{\max}-1} )$. They only search each $(v_r, s_r)$ once, and independently of
the subsequent split rules, hence typically leading to a globally suboptimal solution
$\bar \theta$.

@@

When compared with the described forward stepwise search, a less greedy approach
is to calculate the effects of the split rules deeper down in the
tree. In this way optimal trees can be found for simple problems. However,
split selection at a given node in Equation~\ref{eq:locally} has complexity  $O(PN)$
(if all $P$ variables are numeric/ordered with $N$ distinct values).
Through a global search up to $D-1$ levels (\ref{eq:globally}) the complexity
increases to $O(P^D N^D)$ \citep{papagelis2001bdt}. One conceivable compromise
between these two extremes is to look ahead $d$~steps with $1 < d < D$
\citep[see e.g.,][]{esmeir2007ald}, also yielding a locally optimal tree but
less constrained than that from a 1-step-ahead search.
Another class of algorithms is given by stochastic optimization methods that,
given an initial tree, seek improved solutions through stochastic changes to
the tree structure. Thus, these algorithms try to explore the full parameter
space $\Theta$ but cannot be guaranteed to find the globally optimal solution
but only an approximation thereof. Besides evolutionary algorithms
\citep{koza1991concept}, \textit{Bayesian CART} \citep{denison1998bca} and
\textit{simulated annealing} \citep{sutton1992ict} were used successfully to
solve difficult classification and regression tree problems.
\cite{koza1991concept} first formulated the concept of using evolutionary
algorithms as a stochastic optimization method to build classification and
regression trees. \cite{papagelis2001bdt} presented a classification tree
algorithm and provided results on several datasets from the UCI machine learning
repository \citep{Frank+Asuncion:2010}. Another method for the construction of
classification and regression trees via evolutionary algorithms was introduced by
\cite{gray2008cta} and \cite{fan2005rta}, respectively.
\cite{cantu2003inducing} used an evolutionary algorithm to induce so-called oblique
classification trees.

@@

\textit{Crossover} exchanges, randomly selected, subtrees between two trees. Let
$\tilde{\theta}_{i}^1$ and $\tilde{\theta}_{i}^2$ be the parameters of the trees
that are chosen for crossover. First, an internal node is selected randomly from
each of $\tilde{\theta}_{i}^1$ and $\tilde{\theta}_{i}^2$, which we denote as
$r_1$ and $r_2$. Let $\mathrm{sub}1(\tilde{\theta}_i^j, r_j)$ denote the subtrees
rooted by $r_j$ ($j=1,2$). Thus, $\mathrm{sub}1(\tilde{\theta}_i^j, r_j)$ contains $r_j$, 
and its descendant nodes. Then, the complementary part of $\tilde{\theta}_{i}^j$ can be 
defined as 
$\mathrm{sub}2(\tilde{\theta}_{i}^j, r_j) =  \tilde{\theta}_{i}^j \setminus \mathrm{sub}1(\tilde{\theta}_{i}^j, r_j)$.

After the application of the \textit{crossover} operator, the trees' parameter
spaces are given by $\tilde{\theta}_{i+1}^1= \mathrm{sub}2(\tilde{\theta}_{i}^1, r_1)
\cup \mathrm{sub}1(\tilde{\theta}_{i}^2, r_2) \setminus (v_{m_{21}}, s_{m_{21}}, ...,
v_{m_{2f}}, s_{m_{2f}})$ and $\tilde{\theta}_{i+1}^2=
\mathrm{sub}2(\tilde{\theta}_{i}^2, r_2) \cup \mathrm{sub}1(\tilde{\theta}_{i}^1, r_1)
\setminus (v_{m_{11}}, s_{m_{11}}, ..., v_{m_{1f}}, s_{m_{1f}})$. Here,
$(v_{m_{11}}, s_{m_{11}}, ..., v_{m_{1f}}, s_{m_{1f}}) \subset
\mathrm{sub}1(\tilde{\theta}_i^1, r_1)$ and   
$(v_{m_{21}}, s_{m_{21}}, ...,v_{m_{2f}}, s_{m_{2f}}) \subset \mathrm{sub}1(\tilde{\theta}_i^2, r_2)$ 
are the parameters of the invalid nodes.

@@

First, the data is loaded and the forward-search trees are grown
with and without depth restriction, visualizing the unrestricted trees
in Figure~\ref{fig:example_rpart_ctree}.
%
<<BBBClub-rpart-ctree, echo=TRUE, eval=FALSE>>=
data("BBBClub", package = "evtree")
library("rpart")
rp  <- as.party(rpart(choice ~ ., data = BBBClub, minbucket = 10))
rp2 <- as.party(rpart(choice ~ ., data = BBBClub, minbucket = 10,
  maxdepth = 2))
ct  <- ctree(choice ~ ., data = BBBClub, minbucket = 10, mincrit = 0.99)
ct2 <- ctree(choice ~ ., data = BBBClub, minbucket = 10, mincrit = 0.99,
  maxdepth = 2)
plot(rp)
plot(ct)
@
%
With the objective of building a smaller, but at still accurate tree,
\code{evtree} is constrained to a maximum tree depth of $2$,
see Figure~\ref{fig:example_evtree}.
%
<<BBBClub-evtree, echo=TRUE, eval=FALSE>>=
set.seed(1090)
ev <- evtree(choice ~ ., data = BBBClub, minbucket = 10, maxdepth = 2)
@
%
<<BBBClub-cache, echo=FALSE, eval=TRUE>>=
if(cache & file.exists("BBBClub-trees.rda")) {
load("BBBClub-trees.rda")
} else {
<<BBBClub-rpart-ctree>>
<<BBBClub-evtree>>
if(cache) {
  save(rp, rp2, ct, ct2, ev, file = "BBBClub-trees.rda")
} else {
  if(file.exists("BBBClub-trees.rda")) file.remove("BBBClub-trees.rda")
}
}
@
%
\setkeys{Gin}{width=1\textwidth}
\begin{figure}[p!]
\centering
\begin{minipage}{1\textwidth}
\begin{flushleft} \code{rpart} \end{flushleft}
\centering
\setkeys{Gin}{width=0.75\textwidth}
<<BBBClub-rpart-plot, fig=TRUE, echo=FALSE, width=8.1, height=7>>=
plot(rp)
@
\end{minipage}
\begin{minipage}{1\textwidth}
\begin{flushleft} \code{ctree} \end{flushleft}
\centering
\setkeys{Gin}{width=1\textwidth}
<<BBBClub-ctree-plot, fig=TRUE, echo=FALSE,  width=11.4, height=5.9>>=
plot(ct)
@
\end{minipage}
\caption{\label{fig:example_rpart_ctree} Trees for customer targeting constructed by \code{rpart}
  (upper panel) and \code{ctree} (lower panel). The target variable is the customer's choice of buying the book.
  The variables used for splitting are the number of art books purchased previously (\code{art}), the number
  of months since the first purchase (\code{first}), the frequency of previous purchases at the Bookbinder's Book Club
  (\code{freq}) and the customer's \code{gender}.}
\end{figure}
%
The resulting evolutionary tree is printed below and visualized in
Figure~\ref{fig:example_evtree}.
%
<<BBBClub-evtree-display>>=
plot(ev)
ev
@
%
\setkeys{Gin}{width=1\textwidth}
\begin{figure}[t!]
\begin{flushleft} \code{evtree} \end{flushleft}
\centering
\setkeys{Gin}{width=0.7\textwidth}
<<BBBClub-evtree-plot, fig=TRUE, echo=FALSE, width=7.5, height=4.9>>=
plot(ev)
@
\caption{\label{fig:example_evtree} Tree for customer targeting constructed by \code{evtree}
   The target variable is the customer's choice of buying the book. The variables used for splitting are
   the number of art books purchased previously (\code{art}) and the number of months since the first purchase
    \code{first}.}
\end{figure}

@@

In the first part of the analysis (Section~\ref{comp1}) the algorithms are
compared on 14~benchmark datasets that are publicly available and $3$ real-world
datasets from the Austrian \textit{Diagnosis Related Group (DRG)} system
\citep{LKF2011}. The analysis is based on the evaluation of 250~bootstrap
samples for each of the 20~datasets. The misclassification rate on the
\textit{out-of-bag} \citep{hothorn2005design} samples is used as a measure of
predictive accuracy. Furthermore, the complexity is estimated by the number of
terminal nodes.

@@

In Table~\ref{tab:dataDescription} the benchmark and real-world datasets from
the Austrian DRG system are described. In the Austrian DRG system, resources are 
allocated to hospitals by simple rules mainly regarding the patients' diagnoses, procedures, and age. 
Regression tree analysis is performed to model patient groups with similar resource consumption. 
A more detailed description of the datasets and the application can 
be found in \citet{grubinger2010regression}.

@@

Disadvantages of the \code{evtree} algorithm are computation time and memory requirements. While the smallest of the analyzed datasets, \textit{Glass identification}, only needed approximately 4--6~seconds to fit, larger datasets demanded several minutes. The fit of a model from the largest dataset, \textit{MAGIC gamma telescope}, required approximately 40--50 minutes and a main memory of 400 Mbit. The required resources were measured on an Intel Core 2 Duo with 2.2 GHz and 2 GB RAM using the 64-bit version of Ubuntu 10.10.

Another important issue to be considered is the random nature of evolutionary algorithms. For larger datasets, frequently, considerable different solutions exist that yield a similar or even the same evaluation function value. Therefore, subsequent runs of \code{evtree} can result in very different tree structures. This is not a problem if the tree is intended only for predictive purposes, and it is also not a big issue for many decision and prognosis tasks. Typically, in such applications, the resulting model has to be accurate, compact, and meaningful in its interpretation, but the particular tree structure is of secondary importance. Examples of such applications include the presented marketing case study and the Austrian DRG system. In cases where a model is not meaningful in its interpretation, the possibility of constructing different trees can even be beneficial. However, if the primary goal is to interpret relationships in the data, based on the selected splits, the random nature of the algorithm has to be considered.

@@

In this paper, we presented the \pkg{evtree} package, which implements
classification and regression trees that are grown by an evolutionary
algorithm. The package uses standard \code{print()}, \code{summary()}, and
\code{plot()} functions to display trees and a \code{predict()} function to
predict the class labels of new data from the \pkg{partykit} package. As evolutionary
learning of trees is computationally demanding, most calculations
are conducted in \proglang{C++}. At the moment our algorithm does not support
parallelism. However, we intend to extend \pkg{evtree} to parallel computing.
